One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-20 paper/paper.tex
spell checked

% but also natural in GI.
However, spiral arms are also associated with global gravitational 
instability (GI) of differentially rotating discs
\citep{goldreich65,laughlin96b,laughlin98,nelson98,lodato05,forgan11}. Large-scale
spiral arms can provide significant angular momentum transport
neccessary for mass accretion \citep{lynden-bell72, papaloizou91,balbus99,lodato04}, and
spiral structures due to GI are potentially observable with the Atacama Large 
Millimeter/sub-millimeter Array \citep{cossins10,dipierror14}.

@@

This paper is organised as follows. In \S\ref{model} we list the
governing equations and describe the disc equilibria under
consideration. \S\ref{methods} describes the hydrodynamic codes
employed in this work and several diagnostic measures. We briefly
review basic results from linear theory in \S\ref{wkb} to aid our
interpretation of numerical simulations.  Our main results are presented in 
\S\ref{results2d} for 2D simulations and \S\ref{results3d} for 3D
simulations, and we further analyse them in \S\ref{discussions}. 
We summarise in \S\ref{summary}, with a discussion on
spectulations for future work.

>>>>

% but also natural in GI.
However, spiral arms are also associated with global gravitational 
instability (GI) of differentially rotating discs
\citep{goldreich65,laughlin96b,laughlin98,nelson98,lodato05,forgan11}. Large-scale
spiral arms can provide significant angular momentum transport
necessary for mass accretion \citep{lynden-bell72, papaloizou91,balbus99,lodato04}, and
spiral structures due to GI are potentially observable with the Atacama Large 
Millimeter/sub-millimeter Array \citep{cossins10,dipierror14}.

@@

This paper is organised as follows. In \S\ref{model} we list the
governing equations and describe the disc equilibria under
consideration. \S\ref{methods} describes the hydrodynamic codes
employed in this work and several diagnostic measures. We briefly
review basic results from linear theory in \S\ref{wkb} to aid our
interpretation of numerical simulations.  Our main results are presented in 
\S\ref{results2d} for 2D simulations and \S\ref{results3d} for 3D
simulations, and we further analyse them in \S\ref{discussions}. 
We summarise in \S\ref{summary}, with a discussion on
speculations for future work.

### 2014-11-20 paper/results.tex
spell checked

\subsection{Reference run with random perturbations}
To obtain a picture of the overall disc evolution, we describe a 
fiducial run initialised with $M=10$ in Eq. \ref{randpert}. 
Fig. \ref{fargo_modeamp} plots evolution of the maximum 
non-axisymmetric surface density amplitudes in $R\in[R_{1},R_{2}]$
for $m\in[1,10]$. Snapshots from the simulation are shown in
Fig. \ref{fargo_2d}. 
At early times $t\lesssim100P_0$ the disc 
dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
growth initially and saturate (or decays) after $t=40P_0$. Notice the
low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
possibily due to non-linear interaction of the high-$m$ modes  
\citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
again after $t=70P_0$, and eventually dominates the annulus.

@@

Fig. \ref{2d_fargo_viz} shows the $m=1$ surface density
at the end of this simulation. For this one-armeded spiral
we measure a co-rotation radius and growth rate  
\begin{align*}
  &R_c \simeq 4.4R_0,\\
  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_p. 
\end{align*}
The spiral can be considered as low frequency because its 
pattern speed $\Omega_p \simeq 0.1\Omega_k(R_0)\lesssim 0.3\Omega$ in
$R\in[R_{1},R_{2}]$ (where it has the largest amplitude). Thus, the spiral 
pattern appears nearly stationary. The growth rate $\gamma$ is also slow
relative to the local rotation, although the characteristic growth
time $\gamma^{-1} \simeq 10P_0$ is not very long.  
\begin{figure}
  \includegraphics[width=\linewidth]{figures/polarxy2_dens120_fargo.ps}
  \caption{Cartesian visualisation of the $m=1$ surface density
    structure in the FARGO simulation initialised with only $m=1$
    perturbations. 
    \label{2d_fargo_viz}} 
\end{figure}

@@

In order to track down the origin for the (slow) growth of the
$m=1$ spiral, we invoke angular momentum conservation for linear
perturbations, Eq. \ref{lin_ang_mom_cons}. Assuming fluxes are
negligible at the disc boundaries, we can integrate
Eq. \ref{lin_ang_mom_cons} to give
\begin{align}\label{baroclinic_torque_int}
  \frac{d}{dt}\underbrace{\int_{R_\mathrm{min}}^{R_\mathrm{max}}\jlin
    2\pi R dR}_{\jlintot} 
  =\int_{R\mathrm{min}}^{R\mathrm{max}}T_\mathrm{BG} 2\pi R dR, 
\end{align}
where we recall $T_\mathrm{BG}$ is the torque density due to the imposed
sound-speed profile (Eq. \ref{baroclinic_torque}). We explicitly 
compute both sides of Eq. \ref{baroclinic_torque_int} using
simulation data, and compare them in Fig. \ref{fargo_angmom_ex}. There
is a good match between the two torques, especially at early times
$t\lesssim110P_0$. The average discrepency is $\simeq 5\%$. 
The match is less good later on, when the spiral
amplitude is no longer small ($\Delta\Sigma_1\sim 0.2$ at $t=110P_0$
and $\Delta\Sigma_1\sim 0.4$ by $t=120P_0$) and linear theory becomes
less valid. Fig. \ref{fargo_angmom_ex} confirms that the $m=1$ spiral
wave experiences a negative torque that further reduces its angular
momentum. This is consistent with angular momentum component
measurements (Fig. \ref {2d_angmom}).

>>>>

\subsection{Reference run with random perturbations}
To obtain a picture of the overall disc evolution, we describe a 
fiducial run initialised with $M=10$ in Eq. \ref{randpert}. 
Fig. \ref{fargo_modeamp} plots evolution of the maximum 
non-axisymmetric surface density amplitudes in $R\in[R_{1},R_{2}]$
for $m\in[1,10]$. Snapshots from the simulation are shown in
Fig. \ref{fargo_2d}. 
At early times $t\lesssim100P_0$ the disc 
dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
growth initially and saturate (or decays) after $t=40P_0$. Notice the
low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
possibly due to non-linear interaction of the high-$m$ modes  
\citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
again after $t=70P_0$, and eventually dominates the annulus.

@@

Fig. \ref{2d_fargo_viz} shows the $m=1$ surface density
at the end of this simulation. For this one-armed spiral
we measure a co-rotation radius and growth rate  
\begin{align*}
  &R_c \simeq 4.4R_0,\\
  &\gamma\simeq 0.014\Omega_k(R_0) = 0.13\Omega_p. 
\end{align*}
The spiral can be considered as low frequency because its 
pattern speed $\Omega_p \simeq 0.1\Omega_k(R_0)\lesssim 0.3\Omega$ in
$R\in[R_{1},R_{2}]$ (where it has the largest amplitude). Thus, the spiral 
pattern appears nearly stationary. The growth rate $\gamma$ is also slow
relative to the local rotation, although the characteristic growth
time $\gamma^{-1} \simeq 10P_0$ is not very long.  
\begin{figure}
  \includegraphics[width=\linewidth]{figures/polarxy2_dens120_fargo.ps}
  \caption{Cartesian visualisation of the $m=1$ surface density
    structure in the FARGO simulation initialised with only $m=1$
    perturbations. 
    \label{2d_fargo_viz}} 
\end{figure}

@@

In order to track down the origin for the (slow) growth of the
$m=1$ spiral, we invoke angular momentum conservation for linear
perturbations, Eq. \ref{lin_ang_mom_cons}. Assuming fluxes are
negligible at the disc boundaries, we can integrate
Eq. \ref{lin_ang_mom_cons} to give
\begin{align}\label{baroclinic_torque_int}
  \frac{d}{dt}\underbrace{\int_{R_\mathrm{min}}^{R_\mathrm{max}}\jlin
    2\pi R dR}_{\jlintot} 
  =\int_{R\mathrm{min}}^{R\mathrm{max}}T_\mathrm{BG} 2\pi R dR, 
\end{align}
where we recall $T_\mathrm{BG}$ is the torque density due to the imposed
sound-speed profile (Eq. \ref{baroclinic_torque}). We explicitly 
compute both sides of Eq. \ref{baroclinic_torque_int} using
simulation data, and compare them in Fig. \ref{fargo_angmom_ex}. There
is a good match between the two torques, especially at early times
$t\lesssim110P_0$. The average discrepancy is $\simeq 5\%$. 
The match is less good later on, when the spiral
amplitude is no longer small ($\Delta\Sigma_1\sim 0.2$ at $t=110P_0$
and $\Delta\Sigma_1\sim 0.4$ by $t=120P_0$) and linear theory becomes
less valid. Fig. \ref{fargo_angmom_ex} confirms that the $m=1$ spiral
wave experiences a negative torque that further reduces its angular
momentum. This is consistent with angular momentum component
measurements (Fig. \ref {2d_angmom}).

### 2014-11-20 paper/results_3d.tex
spell checked

The 3D disc has radial size
$[r_\mathrm{min},r_\mathrm{max}]=[0.4,10]R_0$ and vertical extent  
$n_H=2$ scale-heights at $R=R_0$. The resolution is $N_r\times N_\theta\times
N_\phi=256\times32\times512$, or about $4$ cells per
$H$. Because of the reduced resolution 
compared to 2D, we use a smooth perturbation by setting
$\delta = 10^{-3}$ and $M=1$ in Eq. \ref{randpert}. This corresponds
to a single $m=1$ disturbance in $R\in[R_1,R_2]$, which
eventually dominates in 2D simulations.

@@

Fig. \ref{3d_ampmax} plots the evolution of the $m=1$ spiral amplitudes measured
in the ZEUS-MP and PLUTO runs. We also ran simulations
with a strictly isothermal equation of state ($q=0$), which display no
growth compared to that with a temperature gradient.  This confirms
the temperature gradient effect is the same in 3D. 
%why no high m
                                %modes as in 2D for iso disc?
                                %resolution? 
In the ZEUS-MP run we observe disturbances developed near
the inner boundary, which is probably responsible for the initial growth
for $t<70P_0$. This is likely a numerical artifact and effectively seeds
the ZEUS-UP simulation with a larger initial perturbation. Results
from ZEUS-MP are therefore off-set from PLUTO by $\sim50P_0$. However,
once the coherent $m=1$ spiral begins to grow ($t\gtrsim 100P_0$), 
we measure similar growth rates in both codes: 
\begin{align*}
  &\gamma \simeq 0.0073\Omega_k(R_0) \quad\quad \mathrm{PLUTO},\\
  &\gamma \simeq 0.0084\Omega_k(R_0) \quad\quad \varA{ZEUS-MP}.
\end{align*}
Both are smaller than the 2D simulations, because of to the
low resolutions adopted in 3D and/or the small softening length employed in
2D. 
%Fig. \ref{3d_ampmax} also show results from strictly isothermal
%simulations, which display no growth compared to cases with a 
%temperature gradient.

% We also show results from strictly
% isothermal discs ($q=0$). 
%growth rate does not depend on pert amp

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ampmax3d.ps}
  \caption{Evolution of the maximum $m=1$ density component in  $r\in[R_1,R_2]$
    measured in the 3D simulations. Results from discs with a imposed 
    temperature gradient ($q=1$) and a strictly isothermal disc
    ($q=0$) are shown.  
    \label{3d_ampmax}}   
\end{figure}

Visualisations of the 3D simluations are shown in   
Fig. \ref{3d_prelim} for the disc midplane and near the upper disc
boundary. The snapshots are chosen when the one-armed spirals in the two codes
have reached comparable amplitudes. The agreement between the codes,
as well as with the 2D simulations, is satisfactory. The largest
spiral amplitude is found in $R\in[R_1,R_2]$ independent of
height. However, notice the spiral pattern extends into the
non-self-gravitating outer disc ($R>R_2$) at $z\sim 2H$, i.e. 
the disturbance becomes more global away from the midplane.

%Away from the
%disc midplane, a coherent spiral arm is not prominent, but the
%disturbance is radially more global, extending up to $R\sim 4R_0$.

%implications?

\begin{figure}
  \begin{center}
    \subfigure[ZEUS-MP]{
      \includegraphics[scale=0.305,clip=true,trim=0cm 0cm 0cm 0cm]{figures/polarxy2_dens015_z0}  
      \includegraphics[scale=0.305,clip=true,trim=0.8cm 0cm 0cm
      0cm]{figures/polarxy2_dens015_zmax}  
    }
    \subfigure[PLUTO]{
      \includegraphics[scale=0.305,clip=true,trim=0cm 0cm 0cm 0cm]{figures/pdiskxy_023_z0}
      \includegraphics[scale=0.305,clip=true,trim=0.8cm 0cm 0cm
      0cm]{figures/pdiskxy_023_zmax}
    }
  \end{center}
  \caption{Three-dimensional simulations using the ZEUS-MP (top) and 
    PLUTO (bottom) codes. The $m=1$ density component $\Delta\rho_1$ at the midplane (left) and
    two scale-heights above the midplane (right) is shown. Here $\psi
    \equiv \pi/2 - \theta$ is the angular height  
    from the midplane. \label{3d_prelim}}   
\end{figure}

\subsection{Vertical structure}
Fig. \ref{3d_rz} shows the vertical structure of the one-armed
spiral in the PLUTO run. The main spiral is confined to $z < H$. Thus,
a 2D disc model, representing dynamics near the disc midplane, 
should be sufficient to study the instability. Interestingly, however, for $R>2R_0$ the
spirals  increase in amplitude away from the midplane. It remains
small in our disc model ($|\Delta\rho| < 0.1$), but could become 
significant with a larger vertical domain. The figure shows that 
the one-armed spiral can have a global influence away from the
midplane of the self-gravitating annulus.

\begin{figure}
  \includegraphics[scale=0.47,clip=true,trim=0cm 0.79cm 0cm
  0cm]{figures/pdisk_rz_023_sg.ps}  
  \includegraphics[scale=0.47,clip=true,trim=0cm 0cm 0cm
  0.64cm]{figures/pdisk_rz_023_nsg.ps}  
  \caption{Vertical distribution of the $m=1$ density component in the meridional plane from the 
    PLUTO simulation. The slices are taken at the azimuth of   
    $\mathrm{max}[\Delta\rho_1(r,\pi/2,\phi)]$. Arrows correspond to the vector 
    $(v_r/R_0,-v_\theta/rh\sin^2{\theta})$. The top (bottom) panel corresponds
    to the inner (outer) portions of the disc. 
    \label{3d_rz}} 
\end{figure}

Although Fig. \ref{3d_rz} appears to display significant vertical motion,
the three-dimensionality parameter $\Theta < 10^{-2}$
(Eq. \ref{theta}), implying the kinetic energy associated with
vertical motion is insignificant compared to that in horizontal
motion. We find  $|v_z/c_s|\lesssim 
0.2$ for $h=0.05$. Thus, vertical motions are sub-sonic and negligible
compared to horizontal motions, which supports a 2D approximation.

%negligible vertical motion
%unstratified approxi good?

%This again supports the 2D approximation, altthou

%Vertical velocities
%do increase with the growth of the one-armed spiral, 
%Larger amplitudes
%are reached in ZEUS because of larger numerical noise. 
%but we find  $\Theta \lesssim 10^{-2}$.

%\begin{figure}
%  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_vert.ps}
% \caption{Evolution of three-dimensionality $\Theta$, as defined by
%    Eq. \ref{theta}, which corresponds to the
%   average vertical kinetic energy density, scaled by 
%    the average horizontal kinetic energy.\label{3d_vert}} 
%\end{figure}

\subsection{Angular momentum conservation} 
Fig. \ref{3d_angmom} shows the angular momentum evolution in the 3D
runs. ZEUS-MP does not conserve angular momentum very well, but the
variation $|\Delta J/J|< O(10^{-6})$ is small compared 
to the individual components $|\Delta J_{0,1}/J|\sim 
5\times10^{-5}$. The PLUTO run reaches similar values of
$|J_{0,1}|$, but achieves better conservation, with $|\Delta
J/J|=O(10^{-8})$. These plots are again similar to the 2D simulations,
i.e. angular momentum lost by $J_1$ is gained by $J_0$. This confirms 
that the interaction between $J_1$ and $J_0$ operates in 3D and 2D
similarly.

>>>>

The 3D disc has radial size
$[r_\mathrm{min},r_\mathrm{max}]=[0.4,10]R_0$ and vertical extent  
$n_H=2$ scale-heights at $R=R_0$. The resolution is $N_r\times N_\theta\times
N_\phi=256\times32\times512$, or about $4$ cells per
$H$. Because of the reduced resolution 
compared to 2D, we use a smooth perturbation by setting
$\delta = 10^{-3}$ and $M=1$ in Eq. \ref{randpert}. This corresponds
to a single $m=1$ disturbance in $R\in[R_1,R_2]$.
%, which
%eventually dominates in 2D simulations.

@@

Fig. \ref{3d_ampmax} plots the evolution of the $m=1$ spiral amplitudes measured
in the ZEUS-MP and PLUTO runs. We also ran simulations
with a strictly isothermal equation of state ($q=0$), which display no
growth compared to that with a temperature gradient.  This confirms
the temperature gradient effect is the same in 3D. 
%why no high m
                                %modes as in 2D for iso disc?
                                %resolution?

In the ZEUS-MP run, we observed high-$m$ disturbances developed near 
the inner boundary initially, which is likely responsible for the
growth seen at $t<50P_0$. This is a numerical artifact and effectively
seeds the simulation with a larger perturbation. Results
from ZEUS-MP are therefore off-set from PLUTO by $\sim50P_0$. However,
once the coherent $m=1$ spiral begins to grow ($t\gtrsim 100P_0$), 
we measure similar growth rates in both codes: 
\begin{align*}
  &\gamma \simeq 0.0073\Omega_k(R_0) \quad\quad \mathrm{PLUTO},\\
  &\gamma \simeq 0.0085\Omega_k(R_0) \quad\quad \varA{ZEUS-MP}.
\end{align*}
Both are smaller than the 2D simulations, because of the 
lower resolutions adopted in 3D and/or the small softening length
employed in 2D. 
%Rc = 8.6R_0

\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ampmax3d.ps}
  \caption{Evolution of the maximum $m=1$ density component in  $r\in[R_1,R_2]$
    in the 3D simulations. Results from discs with a 
    temperature gradient ($q=1$) and a strictly isothermal disc
    ($q=0$) are shown.  
    \label{3d_ampmax}}   
\end{figure}

Visualisations of the 3D simulations are shown in   
Fig. \ref{3d_prelim} for the disc midplane and near the upper disc
boundary. The snapshots are chosen when the one-armed spirals in the two codes
have reached comparable amplitudes. The agreement between the codes,
as well as with the 2D simulations (Fig. \ref{2d_fargo_viz}), is
satisfactory. The largest spiral amplitude is found in the
self-gravitating region $R\in[R_1,R_2]$, independent of
height. However, notice the spiral pattern extends into 
the non-self-gravitating outer disc ($R>R_2$) at $z\sim 2H$, i.e.    
the disturbance becomes more global away from the midplane.

\begin{figure}
  \begin{center}
    \subfigure[ZEUS-MP]{
      \includegraphics[scale=0.305,clip=true,trim=0cm 0cm 0cm 0cm]{figures/polarxy2_dens015_z0}  
      \includegraphics[scale=0.305,clip=true,trim=0.8cm 0cm 0cm
      0cm]{figures/polarxy2_dens015_zmax}  
    }
    \subfigure[PLUTO]{
      \includegraphics[scale=0.305,clip=true,trim=0cm 0cm 0cm 0cm]{figures/pdiskxy_023_z0}
      \includegraphics[scale=0.305,clip=true,trim=0.8cm 0cm 0cm
      0cm]{figures/pdiskxy_023_zmax}
    }
  \end{center}
  \caption{Three-dimensional simulations using the ZEUS-MP (top) and 
    PLUTO (bottom) codes. The $m=1$ density component $\Delta\rho_1$
    at the midplane (left) and approximately 
    two scale-heights above the midplane (right) is shown. Here $\psi
    \equiv \pi/2 - \theta$ is the angular height  
    from the midplane. \label{3d_prelim}}   
\end{figure}

\subsection{Vertical structure}
Fig. \ref{3d_rz} shows the vertical structure of the one-armed 
spiral in the PLUTO run. The spiral is vertically confined to $z
\lesssim H$ at $R\sim R_0$ (the self-gravitating region). Thus, a 2D
disc model, representing dynamics near the disc midplane, is  
sufficient capture the instability. However, for $R>2R_0$ the spiral amplitude
increases away from the midplane. It remains  
small in our disc model ($|\Delta\rho_1| \lesssim 0.1$), but could become 
significant with a larger vertical domain. This means that 3D 
simulations are necessary to study the effect of the one-armed spiral
on the exterior, non-self-gravitating disc.

\begin{figure}
  \includegraphics[scale=0.47,clip=true,trim=0cm 0.79cm 0cm
  0cm]{figures/pdisk_rz_023_sg.ps}  
  \includegraphics[scale=0.47,clip=true,trim=0cm 0cm 0cm
  0.64cm]{figures/pdisk_rz_023_nsg.ps}  
  \caption{The $m=1$ density component in the meridional plane in the 
    PLUTO simulation. The slices are taken at the azimuth of   
    $\mathrm{max}[\Delta\rho_1(r,\pi/2,\phi)]$. Arrows represent the vector 
    $(v_r/R_0,-v_\theta/rh\sin^2{\theta})$. The top (bottom) panel corresponds
    to the inner (outer) portions of the disc. 
    \label{3d_rz}} 
\end{figure}

Although Fig. \ref{3d_rz} appears to display significant vertical motion,
we measure the three-dimensionality parameter $\Theta < 10^{-2}$
(Eq. \ref{theta}), so vertical motion is insignificant compared to
horizontal motion. This supports a 2D approximation. On the other
hand, we find  $|v_z/c_s|\lesssim 0.2$ which, although sub-sonic, is
not very small.

\subsection{Angular momentum conservation}   
Fig. \ref{3d_angmom} shows the angular momentum evolution in the 3D 
runs during the linear growth of the one-armed spiral. Because the
ZEUS-MP simulation is off-set from PLUTO, the time interval for the
plot was chosen such that the change in the angular momentum
components are comparable in the two codes.

ZEUS-MP does not conserve angular momentum very well, but the
variation in total angular momentum $|\Delta J/J|< O(10^{-6})$ is
small compared to the individual components $|\Delta J_{0,1}/J|\sim
10^{-4}$. %it worsens with time 
PLUTO reaches similar values of $|\Delta J_{0,1}|$, but achieves better
conservation, with $|\Delta J/J|=O(10^{-8})$. These plots are again
similar to the 2D simulations, i.e. angular momentum lost by $J_1$ is
gained by $J_0$. This confirms that the interaction between $J_1$ and
$J_0$ operates in 3D and 2D similarly.

### 2014-11-20 paper/summary.tex
minor updates to tex
spell checked

We used three independent numerical codes --- FARGO in 2D, ZEUS-MP and
PLUTO in 3D --- to verify the growth of  
one-armed spirals in our disc models is due to the imposed temperature
profile and disc structure. The spiral is confined between two
$Q$-barriers owing to a surface density bump. Growth rates increased
with the magnitude of the imposed temperature gradient, and one-armeded
spirals did not develop in strictly isothermal simulations.  
We find no significant vertical 
motion in 3D simulations, which indicates the instability is
two-dimensional.

@@

%works in 2D/3D
%forced gradient feedback 
%neutral mode destabilized      
\subsection{Spectulations and future work}
%balancing heating and cooling 
In the deeply non-linear regime, the one-armed spiral may 
shock and dissipate, depositing negative angular momentum onto
the background disc. The spiral amplitude would saturate by gaining
positive angular momentum. However, if the temperature gradient is
forced, it may be possible to achieve a balance between the gain of negative
angular momentum through the forced temperature gradient, and the gain
of positive angular momentum through shock dissipation. In order to
study this possibility, improved numerical models are needed to ensure 
total angular momentum conservation in long-term simulations.

%numerical simulations to include cooling/heating
A locally isothermal disc represents the limit of infinitely short
cooling (and heating) timescales, so the disc temperature instantly
returns to its inital value. %so we don't go to deeply non-linear
                             %regime 
This assumption can be relaxed by including an energy
equation with thermal relaxation. Preliminary FARGO simulations
indicate a relaxation timescale $t_c < 0.1\Omega_k^{-1}$ is needed for
the one-armed spiral to develop. This will be presented in a follow-up
study. %prelim sims show they are long lasting

>>>>

We used three independent numerical codes --- FARGO in 2D, ZEUS-MP and
PLUTO in 3D --- to verify the growth of  
one-armed spirals in our disc models is due to the imposed temperature
profile and disc structure. The spiral is confined between two
$Q$-barriers owing to a surface density bump. Growth rates increased
with the magnitude of the imposed temperature gradient, and one-armed
spirals did not develop in strictly isothermal simulations.  
We find no significant vertical 
motion in 3D simulations, which indicates the instability is
two-dimensional.

@@

%works in 2D/3D
%forced gradient feedback 
%neutral mode destabilized      
\subsection{Speculations and future work}
%accretion??
%balancing heating and cooling 
In the deeply non-linear regime, the one-armed spiral may 
shock and dissipate, depositing negative angular momentum onto
the background disc. The spiral amplitude would saturate by gaining
positive angular momentum. However, if the temperature gradient is
forced, it may be possible to achieve a balance between the gain of negative
angular momentum through the forced temperature gradient, and the gain
of positive angular momentum through shock dissipation. In order to
study this possibility, improved numerical models are needed to ensure 
total angular momentum conservation in long-term simulations.

%numerical simulations to include cooling/heating
A locally isothermal disc represents the limit of infinitely short
cooling (and heating) timescales, so the disc temperature instantly
returns to its initial value. %so we don't go to deeply non-linear
                             %regime 
This assumption can be relaxed by including an energy
equation with thermal relaxation. Preliminary FARGO simulations
indicate a relaxation timescale $t_c < 0.1\Omega_k^{-1}$ is needed for
the one-armed spiral to develop. This will be presented in a follow-up
study. %prelim sims show they are long lasting
